---
title: "Reproducible Research- Assignment 1"
author: "Micaela"
date: "July 18, 2015"
output: html_document
---

#Loading and preprocessing the data

```{r load in the data needed into your current working directory, echo=TRUE}
temp <- tempfile()
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileURL, temp, method="curl")
data <- read.csv(unzip(temp))
unlink(temp)
(dateDownloaded <- date())
```

#What is mean total number of steps taken per day?
```{r calculate the number of steps taken per day, echo=TRUE}
steps_day <- rowsum(data$steps,group=data$date,na.rm=TRUE)
```

```{r plot number of steps per day,echo=TRUE}
hist(steps_day, main="Steps Taken per Day",xlab="Steps/Day")
```

```{r mean and median steps per day,echo=TRUE}
mean_steps_day <- mean(steps_day,na.rm=TRUE)
median_steps_day <- median(steps_day,na.rm=TRUE)
```
The mean number of steps taken per day is `r round(mean_steps_day,2)`.
The median number of steps taken per day is `r round(median_steps_day,2)`.

#What is the average daily pattern?
```{r calculate and plot the average number of steps per interval,echo=TRUE}
average_interval <- aggregate(data$steps,by=list(data$interval),FUN=mean,na.rm=TRUE)
names(average_interval) <- c("Interval","Average")
plot(average_interval$Interval,average_interval$Average,type="l",
      main="When are the most steps taken on average?",xlab="5 minute interval",
      ylab="Average steps per interval")
max_steps_interval <- average_interval[average_interval$Average==max(average_interval$Average),]
max_steps_interval[,1]
```

#Impute missing data by replacing missing values with the mean number of steps taken for that interval
```{r calculate the number of missing values in the dataset,echo=TRUE}
sum(is.na(data$steps))
```

```{r replace missing values with mean number of steps per interval on other days,echo=TRUE}
data_nomiss <- data
for (i in 1:nrow(data)){
  if (is.na(data[i,"steps"])==TRUE){
    interval <- data[i,"interval"]
    data_nomiss[i,"steps"] <- average_interval[average_interval$Interval==interval,][,2]
    }
  else{
    data_nomiss[i,"steps"] <- data_nomiss[i,"steps"]
  }
}
```

```{r histogram of data with missing data replaced by mean values,echo=TRUE}
steps_day_nomiss <- rowsum(data_nomiss$steps,group=data_nomiss$date)
hist(steps_day_nomiss, main="Steps Taken per Day with Missing Data Imputed",xlab="Steps/Day")
```

```{r mean and median steps per day in the imputed data,echo=TRUE}
mean_steps_day_nomiss <- mean(steps_day_nomiss)
median_steps_day_nomiss <- median(steps_day_nomiss)
```

The mean number of steps taken per day in the imputed data is `r mean_steps_day_nomiss`.
The median number of steps taken per day in the imputed data is `r median_steps_day_nomiss`.

Compare imputed mean and median to original mean and median:
`r round(mean_steps_day_nomiss,digits=2)` and `r round(median_steps_day_nomiss,digits=2)` are larger than `r round(mean_steps_day,digits=2)` and `r round(median_steps_day,digits=2)`.

#Are there differences in activity patterns between weekdays and weekends?
```{r make new variable for weekdays/weekends, echo=TRUE}
data_nomiss$day_of_week <- weekdays(as.Date(data_nomiss$date))
data_nomiss$type_day <- ifelse(data_nomiss$day_of_week == "Saturday" | data_nomiss$day_of_week == "Sunday","Weekend", "Weekday")

data_nomiss$type_day <- as.factor(data_nomiss$type_day)
```

```{r plot steps taken by weekend/weekday, echo=TRUE}
require(lattice)
average_interval_type_day <- aggregate(data_nomiss$steps,by=c(list(data_nomiss$interval),list(data_nomiss$type_day)),FUN=mean)
names(average_interval_type_day) <- c("Interval","Type_Day", "Average")
xyplot(average_interval_type_day$Average~average_interval_type_day$Interval|
         average_interval_type_day$Type_Day,layout=c(1,2),type="l",xlab="Interval",
       ylab="Average Number of Steps")
```
